An Efficient Forecasting Approach for the Reduction of Boundary Effects in Time-Frequency Representations

### 2020-07-02 Notes/algorithm.tex
paper and code

As explained above, the algorithm for the reduction of boundary effects on time-frequency representations relies on two steps. These ones are detailed in the current section.

\subsection{Forecasting}

Let $x:\RR\to\RR$ denote a continuous-time signal. In this work, we consider a finite-length discretization of that one. Thus, the sampled signal $\bx$, whose length is denoted by $N$, is such that
\[
\bx[n] = x\left(\frac{n}{\fs}\right)\ ,\quad \forall n\in\{0,\ldots,N-1\}\ , 
\]
where $\fs$ denotes the sampling frequency.

\paragraph{Notations} 
Let $M$ and $K$ be two integers such that $M<N$ and $K+M<N$. Then, for all $k\in\{0,\ldots,K-1\}$, we extract from $\bx\in\RR^N$ the sub-signal $\bx_k\in\RR^M$ given by:
\begin{equation}
\bx_k = 
\begin{pmatrix}
\bx[N-K+(k-1)-(M-1)] \\
\vdots \\
\bx[N-K+(k-1)]
\end{pmatrix}\ .
\label{eq:xk}
\end{equation} 
These sub-signals are gathered into the matrix $\bX\in\RR^{M\times K}$ such that:
\begin{equation*}
\bX = 
\begin{pmatrix}
\bx_0 & \cdots & \bx_{K-1}
\end{pmatrix}\ .
\end{equation*}
Notice that these sub-signals are overlapping each other. Indeed, $\bx_{k+1}$ is a shifting of $\bx_k$ from one sample. We also consider the matrix $\bY\in\RR^{M\times K}$ given by:
\begin{equation*}
\bY = 
\begin{pmatrix}
\bx_1 & \cdots & \bx_{K}
\end{pmatrix}\ .
\end{equation*}

\paragraph{Dynamical model and forecasting} 
Establishing a dynamical model consists in determining the relation linking $\bY$ to $\bX$, that is finding a function $f$ so that
\[
\bY = f(\bX)\ .
\]
In a general framework, forecasting means estimating the function $f$ from the observed values taken by the signal, in order to predict its future values. For instance, the dynamic mode decomposition~\cite{Schmid10dynamic,Williams15data} allows this by setting very few additional constraints on the regularity of $f$. We will see later, in section~\ref{se:theoretical}, that it is not necessary to consider such a complex dynamic model for the study of the signals of interest to us. That is why we consider here a naive dynamical model, assuming that we have the following relation:
\begin{equation}
\bY = \bA\bX\ ,
\label{eq:dyn.model}
\end{equation}
where $\bA\in\RR^{M\times M}$. We adopt a classical strategy in the study of dynamical systems, that is the linearization of a nonlinear phenomenon. Notice that this linear dynamical model can be written equivalently in function of the sub-signals $\bx_k$, as:
\begin{equation}
\bx_{k+1} = \bA\bx_{k},\ ,\forall k\in\{0,\dots,K-1\}\ .
\end{equation}

The forecasting method consists in estimating the unknown matrix $\bA$. Indeed, let $\tilde\bA$ denotes the estimate of $\bA$, we then obtain the forecasting of the signal at time $\frac{N-1+\ell}{\fs}$ by:
\begin{equation}
\label{eq:prediction}
\tilde \bx [N-1+\ell] = \balpha^{(\ell)}\bx_{K}\ ,
\end{equation}  
where $\balpha^{(\ell)}$ denotes the last row of $\tilde\bA^\ell$, that is to say:
\begin{equation}
\balpha^{(\ell)}=\be_M^T\tilde\bA^\ell\ ,
\label{eq:alpha.l}
\end{equation}
where $\be_M$ is the vector of length $M$ given by $\be_M = \begin{pmatrix} 0 & \cdots & 0 & 1\end{pmatrix}^T$.

\paragraph{Model estimation} To estimate the matrix $\bA$, we basically implement the least square estimator. Thus, we solve the following problem:
\begin{equation}
\label{eq:ls.pb}
\tilde\bA = \arg\min_{\balpha} \cL(\bA)\ ,
\end{equation}
where the loss function $\cL$ is given by:
\[
\cL(\bA) = \|\bY-\bA\bX\|^2 = \sum_{k=0}^{K-1} \|\bx_{k+1}-\bA\bx_k\|^2.
\]
Therefore, solving the problem~\eqref{eq:ls.pb}, \ie~$\nabla \cL(\tilde\bA)=\bzero$, gives the following estimate $\tilde\bA$ of the dynamical model matrix $\bA$:
\begin{align}
\tilde\bA &= \bY\bX^T(\bX\bX^T)^{-1}\ .
\label{eq:lse}
\end{align}

@@

\paragraph{Signal extension} In order to reduce the boundary effects on both "sides" of the time-frequency (or time-scale) representation, we finally construct the extended signal $\tilde\bx\in\RR^{N+2L}$ concatenating the backward prediction $\tilde\bx_{\rm bw}$, the observed signal $\bx$, and the forward prediction $\tilde\bx_{\rm fw}$. We summarize the extension step in Algorithm~\ref{alg:extension}. Notice that we handle the backward estimation using the same strategy than described above, but applying it to the reverse signal $\bx^{\rm r} = \begin{pmatrix} \bx[N-1] & \cdots & \bx[0] \end{pmatrix}^T$.

\begin{algorithm}
\caption{Signal extension. $\tilde\bx = \mathsf{SigExt}(\bx,M,K,L)$}
\label{alg:extension}
\begin{algorithmic}
\STATE {\bf Inputs}: $\bx$, $M$, $K$, $L$
\STATE 
\STATE {\bf Forward forecasting.}
\STATE \quad\textbullet\ LS estimation of the forward matrix $\tilde\bA_{\rm fw}$ via equation~\eqref{eq:lse}.
\STATE \quad\textbullet\ Forward forecasting $\tilde\bx_{\rm bw}$ obtained applying equation~\eqref{eq:prediction} with $\ell\in\{1,\ldots,L\}$.
\STATE 
\STATE {\bf Backward forecasting.}
\STATE \quad\textbullet\ Reverse signal $\bx$ to $\bx^{\rm r}$. 
\STATE \quad\textbullet\ LS estimation of the backward matrix $\tilde\bA_{\rm bw}$ via equation~\eqref{eq:lse} applied to $\bx^{\rm r}$.
\STATE \quad\textbullet\ Reversed backward forecasting $\tilde\bx_{\rm bw}^{\rm r}$ obtained applying equation~\eqref{eq:prediction} to $\bx^r$ with $\ell\in\{1,\ldots,L\}$.
\STATE \quad\textbullet\ Reverse $\tilde\bx_{\rm bw}^{\rm r}$ to obtain the estimate $\tilde\bx_{\rm bw}$.
\STATE 
\STATE {\bf Output}: Extended signal $\tilde\bx = \begin{pmatrix} \tilde\bx_{\rm bw}  & \bx & \tilde\bx_{\rm fw}\end{pmatrix}^T$.
\end{algorithmic}
\end{algorithm}

\subsection{Representation}

Let $\ccF_{N}:\RR^{N}\to\RR^{F\times N}$ generically denotes the time-frequency or time-scale representation we are interested in. It can be, for instance, such as short-time Fourier transform (STFT), the continuous wavelet transform (CWT), the synchrosqueezing transform (SST), or the reassignment (RS). Here, $F$ typically denotes the size of the representation along the frequency axis. Due to the boundary effects, the representation $\ccF_{N}(\bx)$ shows undesired patterns when approaching its edges.For example, the instantaneous frequencies highlighted by the SST can be blurred near that edges. To limit these phenomena, we apply the representation to the estimated extended signal $\tilde\bx$. This strategy moves the boundary effects out of the time interval $I=[0, \frac{N-1}{\fs}]$. Finally, the boundary-effects insensitive representation $\ccF_N^{\mathrm{ext}}:\RR^{N}\to\RR^{F\times N}$ of $\bx$ is given for all $\nu\in\{0,\cdots,F-1\}$, $n\in\{0,\cdots,N-1\}$ by:
\begin{equation}
\ccF_N^{\mathrm{ext}}(\bx)[\nu,n] = \ccF_{N+2L}(\tilde \bx)[\nu,L+n]\ .
\label{eq:restriction}
\end{equation}
This amounts to restricting the representation $\ccF_{N+2L}(\tilde \bx)$ to the original measurement interval of $\bx$. For the sake of simplicity, we denotes the restriction operator by $\cR$, where $\cR:\RR^{F\times (N+2L)}\to\RR^{F\times N}$. Consequently, we have:
\begin{equation*}
\ccF_N^{\mathrm{ext}}(\bx) = \cR\left( \ccF_{N+2L}(\tilde \bx) \right) \ .
\end{equation*}

\subsection{Global algorithm}
Finally, the global procedure we implement to reduce boundary effects on windowing-based representations is summarized by the pseudo-code of Algorithm~\ref{alg:boundary}.

\begin{algorithm}
\caption{Tackling boundary effects. $\bF_\bx = \mathsf{BoundEffRed(}\bx,M,K,L,\ccF)$}
\label{alg:boundary}
\begin{algorithmic}
\STATE {\bf Inputs}: $\bx$, $M$, $K$, $L$, $\ccF$
\STATE 
\STATE {\bf Forecasting step.}
\STATE \quad\textbullet\ Signal extension: $\tilde\bx = \mathrm{SigExt}(\bx)$.
\STATE 
\STATE {\bf Representation step.}
\STATE \quad\textbullet\ Representation evaluation: $\ccF_{N+2L}(\tilde\bx)$.
\STATE \quad\textbullet\ Restriction of $\ccF_{N+2L}(\tilde\bx)$ to the central time interval (see~\eqref{eq:restriction}) to obtain $\bF_\bx=\ccF_N^{\mathrm{ext}}(\bx)$.
\STATE 
\STATE {\bf Output}: Signal representation $\bF_\bx$
\end{algorithmic}
\end{algorithm}

>>>>

As explained above, the algorithm for the reduction of boundary effects on TF representations relies on extending the signal by forecasting it before applying the TF analysis. 
%

%\paragraph{Notations} 
We start with the notation. Let $x:\RR\to\RR$ denote a continuous-time signal. In this work, we consider a finite-length discretization of that one. Thus, the sampled signal $\bx$, whose length is denoted by $N$, is such that
\[
\bx[n] = x\left(\frac{n}{\fs}\right)\ ,\quad \forall n\in\{0,\ldots,N-1\}\ , 
\]
where $\fs$ denotes the sampling frequency. 
Let $M$ and $K$ be two integers such that $M<N$ and $K+M<N$. Then, for all $k\in\{0,\ldots,K-1\}$, we extract from $\bx\in\RR^N$ the sub-signal $\bx_k\in\RR^M$ given by:
\begin{equation}
\bx_k = 
\begin{pmatrix}
\bx[N-K+(k-1)-(M-1)] \\
\vdots \\
\bx[N-K+(k-1)]
\end{pmatrix}\ .
\label{eq:xk}
\end{equation} 
These sub-signals are gathered into the matrix $\bX\in\RR^{M\times K}$ such that:
\begin{equation*}
\bX = 
\begin{pmatrix}
\bx_0 & \cdots & \bx_{K-1}
\end{pmatrix}\ .
\end{equation*}
Notice that these sub-signals are overlapping each other. Indeed, $\bx_{k+1}$ is a shifting of $\bx_k$ from one sample. We also consider the matrix $\bY\in\RR^{M\times K}$ given by:
\begin{equation*}
\bY = 
\begin{pmatrix}
\bx_1 & \cdots & \bx_{K}
\end{pmatrix}\ .
\end{equation*}
The boundary effect reduction algorithm is based on manipulating $\bX$ and $\bY$.

%
The pseudo-code of the proposed algorithm to reduce boundary effects on windowing-based TF representations is shown in the Algorithm~\ref{alg:boundary}. We coined the algorithm {\sf BoundEffRed}. Below, we detail the algorithm, particularly the Signal extension algorithm {\sf SigExt} in Algorithm~\ref{alg:boundary}.

\begin{algorithm}
\caption{Tackling boundary effects. $\bF_\bx = \mbox{\sf BoundEffRed}(\bx,M,K,L,\ccF)$}
\label{alg:boundary}
\begin{algorithmic}
\STATE {\bf Inputs}: $\bx$, $M$, $K$, $L$, $\ccF$
\STATE 
\STATE {\bf Forecasting step.}
\STATE \quad\textbullet\ Signal extension: $\tilde\bx = \mbox{\sf SigExt}(\bx)$.
\STATE 
\STATE {\bf Representation step.}
\STATE \quad\textbullet\ Representation evaluation: $\ccF_{N+2L}(\tilde\bx)$.
\STATE \quad\textbullet\ Restriction of $\ccF_{N+2L}(\tilde\bx)$ to the central time interval (see~\eqref{eq:restriction}) to obtain $\bF_\bx=\ccF_N^{\mathrm{ext}}(\bx)$.
\STATE 
\STATE {\bf Output}: Signal representation $\bF_\bx$
\end{algorithmic}
\end{algorithm}

\subsection{Step 1: Extension by forecasting}

\paragraph{Dynamical model and forecasting} 
Establishing a dynamical model consists in determining the relation linking $\bY$ to $\bX$, that is finding a function $f$ so that
\[
\bY = f(\bX)\ .
\]
In a general framework, forecasting means estimating the function $f$ from the observed values taken by the signal, in order to predict its future values. For instance, the dynamic mode decomposition~\cite{Schmid10dynamic,Williams15data} or other more complicated models \cite{Roberts13Gaussian,DeLivera11forecasting,west2006bayesian,vlachas2018data}, allow this by setting additional constraints on the behavior of $f$. We will see later, in section~\ref{se:theoretical}, that it is not necessary to consider such a complex dynamic model for the study of the oscillatory signals of interest to us. That is why we consider here a naive dynamical model, assuming that we have the following relation:
\begin{equation}
\bY = \bA\bX\ ,
\label{eq:dyn.model}
\end{equation}
where $\bA\in\RR^{M\times M}$. In other words, we adopt a classical strategy in the study of dynamical systems, the {\em linearization} of a nonlinear phenomenon, when the system is sufficient regular. Notice that this linearized dynamical model can be written equivalently in function of the sub-signals $\bx_k$, as:
\begin{equation}
\bx_{k+1} = \bA\bx_{k},\ ,\forall k\in\{0,\dots,K-1\}\ .
\end{equation}

The forecasting method consists in estimating the unknown matrix $\bA$. Indeed, let $\tilde\bA$ denotes the estimate of $\bA$. We then obtain the forecasting of the signal at time $\frac{N-1+\ell}{\fs}$ by:
\begin{equation}
\label{eq:prediction}
\tilde \bx [N-1+\ell] = \balpha^{(\ell)}\bx_{K}\ ,
\end{equation}  
where $\balpha^{(\ell)}$ denotes the last row of $\tilde\bA^\ell$; that is to say,
\begin{equation}
\balpha^{(\ell)}=\be_M^T\tilde\bA^\ell\ ,
\label{eq:alpha.l}
\end{equation}
where $\be_M$ is the unit vector of length $M$ given by $\be_M = \begin{pmatrix} 0 & \cdots & 0 & 1\end{pmatrix}^T$.

\paragraph{Model estimation} To estimate the matrix $\bA$, we consider the simple but numerically efficient least square estimator. That is, we solve the following problem:
\begin{equation}
\label{eq:ls.pb}
\tilde\bA = \arg\min_{\balpha} \cL(\bA)\ ,
\end{equation}
where the loss function $\cL$ is given by:
\[
\cL(\bA) = \|\bY-\bA\bX\|^2 = \sum_{k=0}^{K-1} \|\bx_{k+1}-\bA\bx_k\|^2.
\]
Therefore, solving the problem~\eqref{eq:ls.pb}, \ie~$\nabla \cL(\tilde\bA)=\bzero$, gives the following estimate $\tilde\bA$ of the dynamical model matrix $\bA$:
\begin{align}
\tilde\bA &= \bY\bX^T(\bX\bX^T)^{-1}\ .
\label{eq:lse}
\end{align}

@@

\paragraph{Signal extension} In order to reduce the boundary effects on both ``sides'' of the TF (or time-scale) representation, we finally construct the extended signal $\tilde\bx\in\RR^{N+2L}$ concatenating the backward prediction $\tilde\bx_{\rm bw}$, the observed signal $\bx$, and the forward prediction $\tilde\bx_{\rm fw}$. We summarize the extension step in Algorithm~\ref{alg:extension}. Notice that we handle the backward estimation using the same strategy than described above, but applying it to the reverse signal $\bx^{\rm r} = \begin{pmatrix} \bx[N-1] & \cdots & \bx[0] \end{pmatrix}^T$.

\begin{algorithm}
\caption{Signal extension. $\tilde\bx = \mbox{\sf SigExt}(\bx,M,K,L)$}
\label{alg:extension}
\begin{algorithmic}
\STATE {\bf Inputs}: $\bx$, $M$, $K$, $L$
\STATE 
\STATE {\bf Forward forecasting.}
\STATE \quad\textbullet\ LS estimation of the forward matrix $\tilde\bA_{\rm fw}$ via equation~\eqref{eq:lse}.
\STATE \quad\textbullet\ Forward forecasting $\tilde\bx_{\rm bw}$ obtained applying equation~\eqref{eq:prediction} with $\ell\in\{1,\ldots,L\}$.
\STATE 
\STATE {\bf Backward forecasting.}
\STATE \quad\textbullet\ Reverse signal $\bx$ to $\bx^{\rm r}$. 
\STATE \quad\textbullet\ LS estimation of the backward matrix $\tilde\bA_{\rm bw}$ via equation~\eqref{eq:lse} applied to $\bx^{\rm r}$.
\STATE \quad\textbullet\ Reversed backward forecasting $\tilde\bx_{\rm bw}^{\rm r}$ obtained applying equation~\eqref{eq:prediction} to $\bx^r$ with $\ell\in\{1,\ldots,L\}$.
\STATE \quad\textbullet\ Reverse $\tilde\bx_{\rm bw}^{\rm r}$ to obtain the estimate $\tilde\bx_{\rm bw}$.
\STATE 
\STATE {\bf Output}: Extended signal $\tilde\bx = \begin{pmatrix} \tilde\bx_{\rm bw}  & \bx & \tilde\bx_{\rm fw}\end{pmatrix}^T$.
\end{algorithmic}
\end{algorithm}

\subsection{Step 2: Extended Time-Frequency Representation}

Let $\ccF_{N}:\RR^{N}\to\RR^{F\times N}$ generically denotes the TF or time-scale representation we are interested in, which could be, for instance, STFT, CWT, SST, or RS. Here, $F$ typically denotes the size of the discretization along the frequency axis. Due to the boundary effects, the representation $\ccF_{N}(\bx)$ shows undesired patterns when approaching its edges. %For example, the instantaneous frequencies highlighted by the SST can be blurred near that edges. 
To alleviate the boundary effect, we apply the representation to the estimated extended signal $\tilde\bx$. This strategy moves the boundary effects out of the time interval $I=[0, \frac{N-1}{\fs}]$. Finally, the boundary-effects insensitive representation $\ccF_N^{\mathrm{ext}}:\RR^{N}\to\RR^{F\times N}$ of $\bx$ is given for all $\nu\in\{0,\cdots,F-1\}$, $n\in\{0,\cdots,N-1\}$ by:
\begin{equation}
\ccF_N^{\mathrm{ext}}(\bx)[\nu,n] = \ccF_{N+2L}(\tilde \bx)[\nu,L+n]\ .
\label{eq:restriction}
\end{equation}
This amounts to restricting the representation $\ccF_{N+2L}(\tilde \bx)$ to the original measurement interval of $\bx$. For the sake of simplicity, we denotes the restriction operator by $\cR$, where $\cR:\RR^{F\times (N+2L)}\to\RR^{F\times N}$. Consequently, we have:
\begin{equation*}
\ccF_N^{\mathrm{ext}}(\bx) = \cR\left( \ccF_{N+2L}(\tilde \bx) \right) \ .
\end{equation*}
We call $\ccF_N^{\mathrm{ext}}$ the {\em boundary-free TF representation}.
